Appendix to "Fast approximate Bayesian inference of HIV indicators using PCA adaptive Gauss-Hermite quadrature

### 2023-07-04 src/docs_paper/appendix.Rmd
Table, fixes to bold and format of appendix #55

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  dpi = 320,
  cache = TRUE,
  out.width = "95%",
  fig.align = 'center'
)
```

\newpage

@@

\newpage

@@

Here, we describe the simplified Naomi model [@eaton2021naomi] in more detail.

@@

Consider the most recent national household survey with HIV testing which has taken place in the country of interest.
Let $x \in \{1, \ldots, n\}$ refer to a district located within the Spectrum [@stover2010spectrum] region $R_x$.
Let $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, \text{80+}\}$ be five-year age groups.
As short-hand, we write $a = l$ to refer to the age group with lower bound $l$, e.g. $a = 20$ for $a = \text{20-25}$.
We index the known quantity population size $N_{x, s, a}$ by district, sex and age-band, as well as the following unknown quantities: HIV prevalence $\rho_{x, s, a} \in [0, 1]$, ART coverage $\alpha_{x, s, a} \in [0, 1]$, annual HIV incidence rate $\lambda_{x, s, a} > 0$, and proportion of HIV positive persons recently infected $\kappa_{x, s, a} \in [0, 1]$.
Sometimes data are observed at an aggregate level, rather than the more granular modelled level.
We then use $\{\cdot\}$ to generically refer to a aggregate set over which an observation is made, e.g. $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for the adult age range 15-49.
Let $\sum_{\{x\}}$ be shorthand for $\sum_{x \in \{x\}}$, analogously for $s$ and $a$.

@@

We use structured random effects to enable partial pooling of information across units assessed as being similar, such as those belonging to neighbouring districts or adjacent age-bands.
Let $u$ be a generic random effect, with length $\dim(u)$.
Three types of structured random effects are used in the model.
First, we specify the first order auto-regressive model by $u \sim \text{AR}1(\sigma, \phi)$ such that
\begin{align*}
u_1 &\sim \left( 0, \frac{1}{1 - \rho^2} \right), \\
u_i &= \rho u_{i - 1} + \epsilon_i, \quad i = 2, \ldots, \dim(u)
\end{align*}
where $\epsilon_i \sim \mathcal{N}(0, 1)$ are independent and identically distributed (IID) Gaussian random variables, $\sigma > 0$ is the marginal standard deviation, and $|\rho| < 1$ is the (lag-one) correlation parameter.
Second, we use $u \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with full conditionals
$$
u_i \, | \, u_{-i} \sim \mathcal{N} \left(\frac{\sum_{j: j \sim i} u_j}{n_{\delta i}}, \frac{\sigma^2}{n_{\delta i}}\right),
$$
where $u_{-i}$ is $u$ with the $i$th element removed i.e. $(u_1, \ldots, u_{i - 1}, u_{i + 1}, \ldots, u_{\dim(u)})$, $j \sim i$ if the units $i$ and $j$ are defined as adjacent, $n_{\delta i} = |\{j:j \sim i\}|$ is the total number of adjacent units, and $\sigma > 0$ is the marginal standard deviation.
We follow recommendations of @freni2018note on scaling of precision matrices, disconnected adjacency graph components, and islands.
Third, for the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising], we write $u \sim \text{BYM}2(\sigma, \phi)$, where $u$ is comprised of a spatially structured ICAR component $v$ with proportion $\phi \in (0, 1)$ and spatially unstructured IID component $w$ with proportion $1 - \phi$, both scaled to have generalised variance equal to one, and $\sigma > 0$ is the marginal standard deviation such that
$$
u = \sigma \left( \sqrt\phi \cdot v + \sqrt{1 - \phi} \cdot w \right).
$$

@@

We assume the household survey was run according to a complex survey design where each individual $j$ in the population $U$ has non-zero probability $\pi_j \in (0, 1)$ of appearing in the sample $S \subseteq U$. Suppose we observe an outcome $\theta_j \in \{0, 1\}$ for $j \in S$.
Let $w_j = 1 / \pi_j \times 1 / \omega_j$ be design weights, where $\omega_j$ is a non-response factor, then a survey weighted mean, calculated using the `survey` \textsc{R} package [@lumley2004analysis], is given by
$$
\hat \theta = \frac{\sum_{j \in S} w_j \theta_j}{\sum_{j \in S} w_j}.
$$
We account for survey weighting in the variance via the Kish effective sample size [@kish1965survey]
$$
m^{\theta} = \frac{\left(\sum_{j \in S} w_j\right)^2}{\sum_{j \in S} w_j^2}.
$$
The observed number of indicator cases is then $y^{\theta} = m^{\theta} \cdot \hat \theta$.

@@

We model HIV prevalence $\rho_{x, s, a} \in [0, 1]$ on the logit scale using the linear predictor
\begin{equation}
\text{logit}(\rho_{x, s, a}) = \beta^\rho_0 + \beta_{S}^{\rho, s = \text{M}} + u^\rho_a + u_a^{\rho, s = \text{M}} + u^\rho_x + u_x^{\rho, s = \text{M}} + u_x^{\rho, a < 15} + \eta^\rho_{R_x, s, a}. \label{eq:prev}
\end{equation}
Table \ref{tab:prev} provides a description of the terms included in Equation \ref{eq:prev}.

\begin{table}[h]
\centering
\begin{tabularx}{\textwidth}{llX}
\toprule
Term & Distribution & Description \\ 
\midrule
$\beta^\rho_0$ & $\mathcal{N}(0, 5)$ & Intercept \\
$\beta_{s}^{\rho, s = \text{M}}$ & $\mathcal{N}(0, 5)$ & The difference in logit prevalence for men compared to women \\
$u^\rho_a$ & $\text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ & Age random effects for women \\
$u_a^{\rho, s = \text{M}}$ & $\text{AR}1(\sigma_{AS}^\rho, \phi_{AS}^\rho)$ & Age random effects for the difference in logit prevalence for men compared to women age $a$ \\
$u^\rho_x$ & $\text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ & Spatial random effects for women \\
$u_x^{\rho, s = \text{M}}$ & $\text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ & Spatial random effects for the difference in logit prevalence for men compared to women in district $x$  \\
$u_x^{\rho, a < 15}$ & $\text{ICAR}(\sigma_{XA}^\rho)$ & Spatial random effects for the ratio of paediatric prevalence to adult women prevalence \\
$\eta^\rho_{R_x, s, a}$ & $-$ & Fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available \\
\bottomrule
\end{tabularx}
\caption{Terms included in the linear predictor for HIV prevalence (Equation \ref{eq:prev}).}
\label{tab:prev}
\end{table}

The two BYM2 random effects $u^\rho_x$ and $u_x^{\rho, s = \text{M}}$ are comprised of the following unit scaled spatially structured $\{v_x^{\rho}, v_x^{\rho, s = \text{M}}\}$ and spatially unstructured $\{w_x^{\rho}, w_x^{\rho, s = \text{M}}\}$ components, respectively
\begin{align*}
u_x^{\rho} &= \sigma_{X}^\rho \left( \sqrt\phi_{X}^\rho \cdot v_x^{\rho} + \sqrt{1 - \phi_{X}^\rho} \cdot w_x^{\rho} \right), \\
u_x^{\rho, s = \text{M}} &= \sigma_{XS}^\rho \left( \sqrt\phi_{XS}^\rho \cdot v_x^{\rho, s = \text{M}} + \sqrt{1 - \phi_{XS}^\rho} \cdot w_x^{\rho, s = \text{M}} \right).
\end{align*}
We use half-normal priors for the standard deviation terms
$$
\{\sigma_A^\rho, \sigma_{AS}^\rho, \sigma_X^\rho, \sigma_{XS}^\rho, \sigma_{XA}^\rho\} \sim \mathcal{N}^{+}(0, 2.5),
$$
uniform priors for the AR1 correlation parameters
$$
\{\phi_A^\rho, \phi_{AS}^\rho\} \sim \mathcal{U}(-1, 1),
$$
and beta priors for the BYM2 proportion parameters
$$
\{\phi_X^\rho, \phi_{XS}^\rho\} \sim \text{Beta}(0.5, 0.5).
$$

@@

We model ART coverage $\alpha_{x, s, a} \in [0, 1]$ on the logit scale using the linear predictor
$$
\text{logit}(\alpha_{x, s, a}) = \beta^\alpha_0 + \beta_{S}^{\alpha, s = \text{M}} + u^\alpha_a + u_a^{\alpha, s = \text{M}} + u^\alpha_x + u_x^{\alpha, s = \text{M}} + u_x^{\alpha, a < 15} + \eta^\alpha_{R_x, s, a} 
$$
with terms and priors analogous to the HIV prevalence process model in Section \ref{sec:hiv-prev} above.

@@

We model HIV incidence rate $\lambda_{x, s, a} > 0$ on the log scale using the linear predictor
\begin{equation}
\log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a}^\lambda. \label{eq:inc}
\end{equation}
Table \ref{tab:inc} provides a description of the terms included in Equation \ref{eq:inc}.

@@

\begin{table}[h]
\centering
\begin{tabularx}{\textwidth}{llX}
\toprule
Term & Distribution & Description \\ 
\midrule
$\beta^\lambda_0$ & $\mathcal{N}(0, 5)$ & Intercept term proportional to the average HIV transmission rate for untreated HIV positive adults \\
$\beta_S^{\lambda, s = \text{M}}$ & $\mathcal{N}(0, 5)$ & The log incidence rate ratio for men compared to women \\
$\rho_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a}}$ & $-$ & The HIV prevalence among adults 15-49 calculated by aggregating age-specific HIV prevalences \\
$\alpha_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}$ & $-$ & The ART coverage among adults 15-49 calculated by aggregating age-specific ART coverages \\
$\omega = 0.7$ & $-$ & Average reduction in HIV transmission rate per increase in population ART coverage fixed based on inputs to the Estimation and Projection Package (EPP) model \\
$u_x^\lambda$ & $\mathcal{N}(0, \sigma^\lambda)$  & IID spatial random effects with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ \\
$\eta^\lambda_{R_x, s, a}$ & $-$ & Fixed log incidence rate ratios by sex and age group calculated from Spectrum model output \\
\bottomrule
\end{tabularx}
\caption{Terms included in the linear predictor for HIV incidence (Equation \ref{eq:inc}). Note that the only source age structure for this part of the model are $\eta^\lambda_{R_x, s, a}$. As Spectrum assumes that there are no new infections in children aged 5-9 or 10-14, or adults aged over 80, the posterior over new infections in these age groups is exactly zero, by definition. We remove these identically zero posteriors from any later inference comparisons.}
\label{tab:inc}
\end{table}

We model the proportion recently infected among HIV positive persons $\kappa_{x, s, a} \in [0, 1]$ as
$$
\kappa_{x, s, a} = 1 - \exp \left(- \lambda_{x, s, a} \cdot \frac{1 - \rho_{x, s, a}}{\rho_{x, s, a}} \cdot (\Omega_T - \beta_T ) - \beta_T \right),
$$
where $\Omega_T \sim \mathcal{N}(\Omega_{T_0}, \sigma^{\Omega_T})$ is the mean duration of recent infection, and $\beta_T \sim \mathcal{N}^{+}(\beta_{T_0}, \sigma^{\beta_T})$ is the false recent ratio.
We use an informative prior for $\Omega_T$ based on the characteristics of the recent infection testing algorithm.
For PHIA surveys this is $\Omega_{T_0} = 130 \text{ days}$ and $\sigma^{\Omega_T} = 6.12 \text{ days}$, and further we assume there is no false recency, such that $\beta_{T_0} = 0.0$ and $\sigma^{\beta_T} = 0.0$.

@@

HIV prevalence $\rho_{x, a}^\text{ANC}$ and ART coverage $\alpha_{x, a}^\text{ANC}$ among pregnant women are modelled as being offset on the logit scale from the corresponding district-age indicators $\rho_{x, F, a}$ and $\alpha_{x, F, a}$ according to
\begin{align*}
\text{logit}(\rho_{x, a}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a}) + \beta^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + \eta_{R_x, a}^{\rho^{\text{ANC}}}, \\
\text{logit}(\alpha_{x, a}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a}) + \beta^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}} + \eta_{R_x, a}^{\alpha^{\text{ANC}}},
\end{align*}
where, for $\theta \in \{\rho, \alpha\}$, $\beta^{\theta^{\text{ANC}}} \sim \mathcal{N}(0, 5)$ are the average differences between population and ANC outcomes, $u_x^{\theta^{\text{ANC}}} \sim \mathcal{N}(0, \sigma_X^{\theta^{\text{ANC}}})$ are IID district random effects with $\sigma_X^{\theta^{\text{ANC}}} \sim \mathcal{N}^+(0, 1)$, and $\eta_{R_x, a}^{\theta^{\text{ANC}}}$ for are offsets for the log fertility rate ratios for HIV positive women compared to HIV negative women and for women on ART to HIV positive women not on ART, calculated from Spectrum model outputs for region $R_x$.

In the full Naomi model, for adult women 15-49 the number of ANC clients $\Psi_{x, a} > 0$ are modelled as
$$
\log (\Psi_{x, a}) = \log (N_{x, \text{F}, a}) + \psi_{R_x, a} + \beta^\psi + u_x^\psi,
$$
where $N_{x, \text{F}, a}$ are the female population sizes, $\psi_{R_x, a}$ are fixed age-sex fertility ratios in Spectrum region $R_x$, $\beta^\psi$ are log rate ratios for the number of ANC clients relative to the predicted fertility, and $u_x^\psi \sim \mathcal{N}(0, \sigma^\psi)$ are district random effects.
Here we fix $\beta^\psi = u_x^\psi = 0$ such that $\Psi_{x, a}$ are simply constants.

@@

1. If survey data on HIV prevalence or ART coverage by age and sex are not available then we set $u_a^\theta = 0$ and $u_{a, s = \text{M}}^\theta = 0$ and use the average age-sex pattern of from the Spectrum offset $\eta_{R_x, s, a}^\theta$. For the Malawi example considered in the main text HIV prevalence and ART coverage data are not available for those aged 65+. As a result, there are $|\{\text{0-4}, \ldots, \text{50-54}\}| = 13$ age groups included for the age random effects.
2. If no ART data, either survey or ART programme, are available but data on ART coverage among ANC clients are available, the level of ART coverage is not identifiable, but spatial variation is identifiable. In this instance, overall ART coverage is determined by the Spectrum offset, and only area random effects are estimated such that $\text{logit} \left(\alpha_{x, s, a} \right) = u_x^\alpha + \eta_{R_x, s, a}^\alpha$.
3. If survey data on recent HIV infection are not included in the model, then $\beta_0^\lambda = \beta_S^{\lambda, s = \text{M}} = u_x^\lambda = 0$. The sex ratio for HIV incidence is determined by the sex incidence rate ratio from Spectrum in the same years and the incidence rate in all districts is modelled assuming the same average HIV transmission rate for untreated adults, but varies according to district estimates of HIV prevalence and ART coverage.

@@

\texttt{u\_rho\_x} & $w^\rho_x$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
\texttt{us\_rho\_x} & $v^\rho_x$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
\texttt{u\_rho\_xs} & $w_x^{\rho, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
\texttt{us\_rho\_xs} & $v_x^{\rho, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
\texttt{u\_rho\_a} & $u^\rho_a$ & Latent & 10 & $\mathbb{R}^{10}$ & \cmark & & \\

>>>>

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  dpi = 320,
  cache = TRUE,
  out.width = "95%",
  fig.align = "center",
  fig.pos = "!h",
  out.extra = ""
)

knitr::opts_chunk$set()
```

\clearpage

@@

\clearpage

@@

In this section we describe the simplified Naomi model [@eaton2021naomi] in more complete mathematical detail than the main text.

@@

Consider the most recent national household survey with HIV testing which has taken place in the country of interest.
Let $x \in \{1, \ldots, n\}$ refer to a district located within the Spectrum [@stover2010spectrum] region $R_x$.
Let $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, \text{80+}\}$ be five-year age groups.
As short-hand, we write $a = l$ to refer to the age group with lower bound $l$, e.g. $a = 20$ for $a = \text{20-25}$.

We index the known quantity population size $N_{x, s, a}$ by district, sex and age-band, as well as the following unknown quantities:

* HIV prevalence $\rho_{x, s, a} \in [0, 1]$,
* ART coverage $\alpha_{x, s, a} \in [0, 1]$,
* annual HIV incidence rate $\lambda_{x, s, a} > 0$, and
* proportion of HIV positive persons recently infected $\kappa_{x, s, a} \in [0, 1]$.

Sometimes data are observed at an aggregate level, rather than the more granular modelled level.
We use $\{\cdot\}$ to generically refer to a aggregate set over which an observation is made, e.g. $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for the adult age range 15-49.
The notation $\sum_{\{x\}}$ is used as shorthand for $\sum_{x \in \{x\}}$, and analogously for $s$ and $a$.

@@

We use structured random effects to enable partial pooling of information across units assessed as being similar, such as those belonging to neighbouring districts or adjacent age-bands.
Let $\bu$ be a generic random effect, with length $\dim(\bu) > 1$.
Three types of structured random effects are used in the model:

1. We specify the first order auto-regressive model by $\bu \sim \text{AR}1(\sigma, \phi)$ such that
\begin{align}
u_1 &\sim \left( 0, \frac{1}{1 - \rho^2} \right), \\
u_i &= \rho u_{i - 1} + \epsilon_i, \quad i = 2, \ldots, \dim(\bu)
\end{align}
where $\epsilon_i \sim \mathcal{N}(0, 1)$ are independent and identically distributed (IID) Gaussian random variables, $\sigma > 0$ is the marginal standard deviation, and $|\rho| < 1$ is the (lag-one) correlation parameter.
2. We use $\bu \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with full conditionals
\begin{equation}
u_i \, | \, \bu_{-i} \sim \mathcal{N} \left(\frac{\sum_{j: j \sim i} u_j}{n_{\delta i}}, \frac{\sigma^2}{n_{\delta i}}\right),
\end{equation}
where $\bu_{-i}$ is $\bu$ with the $i$th element removed i.e. $(u_1, \ldots, u_{i - 1}, u_{i + 1}, \ldots, u_{\dim(\bu)})$, $j \sim i$ if the units $i$ and $j$ are defined as adjacent, $n_{\delta i} = |\{j:j \sim i\}|$ is the total number of adjacent units, and $\sigma > 0$ is the marginal standard deviation.
We follow recommendations of @freni2018note on scaling of precision matrices, disconnected adjacency graph components, and islands.
3. For the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising], we write $\bu \sim \text{BYM}2(\sigma, \phi)$, where $\bu$ is comprised of a spatially structured ICAR component $\bv$ with proportion $\phi \in (0, 1)$ and spatially unstructured IID component $\w$ with proportion $1 - \phi$, both scaled to have generalised variance equal to one, and $\sigma > 0$ is the marginal standard deviation such that
\begin{equation}
\bu = \sigma \left( \sqrt\phi \cdot \bv + \sqrt{1 - \phi} \cdot \w \right).
\end{equation}

@@

We assume the household survey was run according to a complex survey design where each individual $j$ in the population $U$ has non-zero probability $\pi_j \in (0, 1)$ of appearing in the sample $S \subseteq U$. Suppose we observe an outcome $\theta_j \in \{0, 1\}$ for $j \in S$.
Let $w_j = 1 / \pi_j \times 1 / \omega_j$ be design weights, where $\omega_j$ is a non-response factor, then a survey weighted mean is given by
\begin{equation}
\hat \theta = \frac{\sum_{j \in S} w_j \theta_j}{\sum_{j \in S} w_j}.
\end{equation}
We account for survey weighting in the variance via the Kish effective sample size [@kish1965survey]
\begin{equation}
m^{\theta} = \frac{\left(\sum_{j \in S} w_j\right)^2}{\sum_{j \in S} w_j^2}.
\end{equation}
The observed number of indicator cases is then $y^{\theta} = m^{\theta} \cdot \hat \theta$.
To make these calculations we used the `survey` \textsc{R} package [@lumley2004analysis]

@@

We model HIV prevalence $\rho_{x, s, a} \in [0, 1]$ on the logit scale using the linear predictor
\begin{equation}
\text{logit}(\rho_{x, s, a}) = \beta^\rho_0 + \beta_{S}^{\rho, s = \text{M}} + \bu^\rho_a + \bu_a^{\rho, s = \text{M}} + \bu^\rho_x + \bu_x^{\rho, s = \text{M}} + \bu_x^{\rho, a < 15} + \bmeta^\rho_{R_x, s, a}. \label{eq:prev}
\end{equation}
Table \ref{tab:prev} provides a description of the terms included in Equation \ref{eq:prev}.

\begin{table}[h]
\centering
\begin{tabularx}{\textwidth}{llX}
\toprule
Term & Distribution & Description \\ 
\midrule
$\beta^\rho_0$ & $\mathcal{N}(0, 5)$ & Intercept \\
$\beta_{s}^{\rho, s = \text{M}}$ & $\mathcal{N}(0, 5)$ & The difference in logit prevalence for men compared to women \\
$\bu^\rho_a$ & $\text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ & Age random effects for women \\
$\bu_a^{\rho, s = \text{M}}$ & $\text{AR}1(\sigma_{AS}^\rho, \phi_{AS}^\rho)$ & Age random effects for the difference in logit prevalence for men compared to women age $a$ \\
$\bu^\rho_x$ & $\text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ & Spatial random effects for women \\
$\bu_x^{\rho, s = \text{M}}$ & $\text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ & Spatial random effects for the difference in logit prevalence for men compared to women in district $x$  \\
$\bu_x^{\rho, a < 15}$ & $\text{ICAR}(\sigma_{XA}^\rho)$ & Spatial random effects for the ratio of paediatric prevalence to adult women prevalence \\
$\bmeta^\rho_{R_x, s, a}$ & $-$ & Fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available \\
\bottomrule
\end{tabularx}
\caption{Terms included in the linear predictor for HIV prevalence (Equation \ref{eq:prev}).}
\label{tab:prev}
\end{table}

The two BYM2 random effects $\bu^\rho_x$ and $\bu_x^{\rho, s = \text{M}}$ are comprised of the following unit scaled spatially structured $\{\bv_x^{\rho}, \bv_x^{\rho, s = \text{M}}\}$ and spatially unstructured $\{\w_x^{\rho}, \w_x^{\rho, s = \text{M}}\}$ components, respectively
\begin{align}
\bu_x^{\rho} &= \sigma_{X}^\rho \left( \sqrt\phi_{X}^\rho \cdot \bv_x^{\rho} + \sqrt{1 - \phi_{X}^\rho} \cdot \w_x^{\rho} \right), \\
\bu_x^{\rho, s = \text{M}} &= \sigma_{XS}^\rho \left( \sqrt\phi_{XS}^\rho \cdot \bv_x^{\rho, s = \text{M}} + \sqrt{1 - \phi_{XS}^\rho} \cdot \w_x^{\rho, s = \text{M}} \right).
\end{align}
We use half-normal priors for the standard deviation terms
\begin{equation}
\{\sigma_A^\rho, \sigma_{AS}^\rho, \sigma_X^\rho, \sigma_{XS}^\rho, \sigma_{XA}^\rho\} \sim \mathcal{N}^{+}(0, 2.5),
\end{equation}
uniform priors for the AR1 correlation parameters
\begin{equation}
\{\phi_A^\rho, \phi_{AS}^\rho\} \sim \mathcal{U}(-1, 1),
\end{equation}
and beta priors for the BYM2 proportion parameters
\begin{equation}
\{\phi_X^\rho, \phi_{XS}^\rho\} \sim \text{Beta}(0.5, 0.5).
\end{equation}

@@

We model ART coverage $\alpha_{x, s, a} \in [0, 1]$ on the logit scale using the linear predictor
\begin{equation}
\text{logit}(\alpha_{x, s, a}) = \beta^\alpha_0 + \beta_{S}^{\alpha, s = \text{M}} + \bu^\alpha_a + \bu_a^{\alpha, s = \text{M}} + \bu^\alpha_x + \bu_x^{\alpha, s = \text{M}} + \bu_x^{\alpha, a < 15} + \bmeta^\alpha_{R_x, s, a} 
\end{equation}
with terms and priors analogous to the HIV prevalence process model in Section \ref{sec:hiv-prev} above.

@@

We model HIV incidence rate $\lambda_{x, s, a} > 0$ on the log scale using the linear predictor
\begin{equation}
\log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + \bu_x^\lambda + \bmeta_{R_x, s, a}^\lambda. \label{eq:inc}
\end{equation}
Table \ref{tab:inc} provides a description of the terms included in Equation \ref{eq:inc}.

@@

\begin{table}[h]
\centering
\begin{tabularx}{\textwidth}{llX}
\toprule
Term & Distribution & Description \\ 
\midrule
$\beta^\lambda_0$ & $\mathcal{N}(0, 5)$ & Intercept term proportional to the average HIV transmission rate for untreated HIV positive adults \\
$\beta_S^{\lambda, s = \text{M}}$ & $\mathcal{N}(0, 5)$ & The log incidence rate ratio for men compared to women \\
$\rho_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a}}$ & $-$ & The HIV prevalence among adults 15-49 calculated by aggregating age-specific HIV prevalences \\
$\alpha_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}$ & $-$ & The ART coverage among adults 15-49 calculated by aggregating age-specific ART coverages \\
$\omega = 0.7$ & $-$ & Average reduction in HIV transmission rate per increase in population ART coverage fixed based on inputs to the Estimation and Projection Package (EPP) model \\
$\bu_x^\lambda$ & $\mathcal{N}(0, \sigma^\lambda)$  & IID spatial random effects with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ \\
$\bmeta^\lambda_{R_x, s, a}$ & $-$ & Fixed log incidence rate ratios by sex and age group calculated from Spectrum model output \\
\bottomrule
\end{tabularx}
\caption{Terms included in the linear predictor for HIV incidence (Equation \ref{eq:inc}). Note that the only source age structure for this part of the model are $\bmeta^\lambda_{R_x, s, a}$. As Spectrum assumes that there are no new infections in children aged 5-9 or 10-14, or adults aged over 80, the posterior over new infections in these age groups is exactly zero, by definition. We remove these identically zero posteriors from any later inference comparisons.}
\label{tab:inc}
\end{table}

We model the proportion recently infected among HIV positive persons $\kappa_{x, s, a} \in [0, 1]$ as
\begin{equation}
\kappa_{x, s, a} = 1 - \exp \left(- \lambda_{x, s, a} \cdot \frac{1 - \rho_{x, s, a}}{\rho_{x, s, a}} \cdot (\Omega_T - \beta_T ) - \beta_T \right),
\end{equation}
where $\Omega_T \sim \mathcal{N}(\Omega_{T_0}, \sigma^{\Omega_T})$ is the mean duration of recent infection, and $\beta_T \sim \mathcal{N}^{+}(\beta_{T_0}, \sigma^{\beta_T})$ is the false recent ratio.
We use an informative prior for $\Omega_T$ based on the characteristics of the recent infection testing algorithm.
For PHIA surveys this is $\Omega_{T_0} = 130 \text{ days}$ and $\sigma^{\Omega_T} = 6.12 \text{ days}$, and further there is assumed to be no false recency, such that $\beta_{T_0} = 0.0$ and $\sigma^{\beta_T} = 0.0$.

@@

HIV prevalence $\rho_{x, a}^\text{ANC}$ and ART coverage $\alpha_{x, a}^\text{ANC}$ among pregnant women are modelled as being offset on the logit scale from the corresponding district-age indicators $\rho_{x, F, a}$ and $\alpha_{x, F, a}$ according to
\begin{align}
\text{logit}(\rho_{x, a}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a}) + \beta^{\rho^{\text{ANC}}} + \bu_x^{\rho^{\text{ANC}}} + \bmeta_{R_x, a}^{\rho^{\text{ANC}}}, \label{eq:anc1} \\
\text{logit}(\alpha_{x, a}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a}) + \beta^{\alpha^{\text{ANC}}} + \bu_x^{\alpha^{\text{ANC}}} + \bmeta_{R_x, a}^{\alpha^{\text{ANC}}} \label{eq:anc2}.
\end{align}
Table \ref{tab:anc} provides a description of the terms included in Equation \ref{eq:anc1} and Equation \ref{eq:anc2}.

\begin{table}[h]
\centering
\begin{tabularx}{\textwidth}{llX}
\toprule
Term & Distribution & Description \\ 
\midrule
$\beta^{\theta^{\text{ANC}}}$ & $\mathcal{N}(0, 5)$ & Intercept giving the average difference between population and ANC outcomes \\
$\bu_x^{\theta^{\text{ANC}}}$ & $\mathcal{N}(0, \sigma_X^{\theta^{\text{ANC}}})$ & IID district random effects with $\sigma_X^{\theta^{\text{ANC}}} \sim \mathcal{N}^+(0, 1)$ \\ 
$\bmeta_{R_x, a}^{\theta^{\text{ANC}}}$ & $-$ & Offsets for the log fertility rate ratios for HIV positive women compared to HIV negative women and for women on ART to HIV positive women not on ART, calculated from Spectrum model outputs for region $R_x$ \\
\bottomrule
\end{tabularx}
\caption{Terms included in the linear predictors for ANC testing, with $\theta \in \{\rho, \alpha\}$ (Equation \ref{eq:prev}).}
\label{tab:anc}
\end{table}

In the full Naomi model, for adult women 15-49 the number of ANC clients $\Psi_{x, a} > 0$ are modelled as
$$
\log (\Psi_{x, a}) = \log (N_{x, \text{F}, a}) + \psi_{R_x, a} + \beta^\psi + \bu_x^\psi,
$$
where $N_{x, \text{F}, a}$ are the female population sizes, $\psi_{R_x, a}$ are fixed age-sex fertility ratios in Spectrum region $R_x$, $\beta^\psi$ are log rate ratios for the number of ANC clients relative to the predicted fertility, and $\bu_x^\psi \sim \mathcal{N}(0, \sigma^\psi)$ are district random effects.
Here we fix $\beta^\psi = 0$ and $\bu_x^\psi = \mathbf{0}$ such that $\Psi_{x, a}$ are simply constants.

@@

1. If survey data on HIV prevalence or ART coverage by age and sex are not available then we set $\bu_a^\theta = 0$ and $\bu_{a, s = \text{M}}^\theta = 0$ and use the average age-sex pattern of from the Spectrum offset $\bmeta_{R_x, s, a}^\theta$. For the Malawi example considered in the main text HIV prevalence and ART coverage data are not available for those aged 65+. As a result, there are $|\{\text{0-4}, \ldots, \text{50-54}\}| = 13$ age groups included for the age random effects.
2. If no ART data, either survey or ART programme, are available but data on ART coverage among ANC clients are available, the level of ART coverage is not identifiable, but spatial variation is identifiable. In this instance, overall ART coverage is determined by the Spectrum offset, and only area random effects are estimated such that $\text{logit} \left(\alpha_{x, s, a} \right) = \bu_x^\alpha + \bmeta_{R_x, s, a}^\alpha$.
3. If survey data on recent HIV infection are not included in the model, then $\beta_0^\lambda = \beta_S^{\lambda, s = \text{M}} = 0$ and $\bu_x^\lambda = \mathbf{0}$. The sex ratio for HIV incidence is determined by the sex incidence rate ratio from Spectrum in the same years and the incidence rate in all districts is modelled assuming the same average HIV transmission rate for untreated adults, but varies according to district estimates of HIV prevalence and ART coverage.

@@

\texttt{u\_rho\_x} & $\w^\rho_x$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
\texttt{us\_rho\_x} & $\bv^\rho_x$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
\texttt{u\_rho\_xs} & $\w_x^{\rho, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
\texttt{us\_rho\_xs} & $\bv_x^{\rho, s = \text{M}}$ & Latent & $n$ & $\mathbb{R}^{n}$ & \cmark & & \\
\texttt{u\_rho\_a} & $\bu^\rho_a$ & Latent & 10 & $\mathbb{R}^{10}$ & \cmark & & \\
